Approximation as Differentiation

### 2025-06-28 main/language.tex
Drop inductive types from syntax, in favour of lists.
Kinding not needed if we only have lists.
Separate figure for parameterised types.

% \input{fig/positivity} -- follow nunes2023 instead
\input{fig/syntax}
\input{fig/typing}

@@

\input{fig/mu-polynomial}
\input{fig/default-semantics}

\begin{definition}[$\mu$-polynomial]
Suppose $\cat{C}$ a category with finite coproducts $(\coprod, 0)$ and finite products $(\times, 1)$. Define
$\muPoly_{\cat{C}}$ to be the smallest subcategory of $\Cat$ generated inductively by the rules in
\figref{mu-polynomial}.
\end{definition}

For any $\mu$-polynomial endofunctor $F: \cat{C} \to \cat{C}$ with initial algebra $\mu F$, write $\inMap_F$
for the structure map $F\mu F \to \mu F$ and write $\cata_f$ for the unique morphism $\mu F \to X$ from the
initial $F$-algebra to any $F$-algebra $(X, f: FX \to X)$.

To give the semantics for the language defined in \figrefTwo{syntax}{typing}, we fix a bicartesian closed
category $\Sem$ with finite products $(\times, 1)$, finite coproducts $(\coprod, 0)$, exponentials
$\internalHom{X}{Y}$, evaluation morphisms $\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$, plus the
following additional structure:
\begin{enumerate}
\item a strong monad $\mathcal{L}$ (lifting) with unit $\eta_X: X \to \mathcal{L}(X)$, multiplication $\mu_X:
\mathcal{L}^2(X) \to \mathcal{L}(X)$ and strength $\mathsf{st}_{X,Y}: X \times \mathcal{L}(Y) \to
\mathcal{L}(X \times Y)$;
\item $\mu$-polynomial endofunctors given by $\muPoly_{\Sem}$;
\item for each primitive type $\rho \in \PrimTy$, an object $\sem{\rho}$;
\item for each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\phi}:
\sem{\rho_1} \times \ldots \times \sem{\rho_n} \to \sem{\rho}$.
\end{enumerate}

\figref{default-semantics:types} gives the interpretation of types $\Delta \vdash \tau$ with $\length{\Delta}
= n$ as functors $\Sem^n \to \Sem$, or as objects of $\Sem$ in the case $n = 0$.
\figref{default-semantics:terms} gives the interpretation of terms $\Gamma \vdash t: \tau$ as morphisms
$\sem{\Gamma} \to \sem{\tau}$.

\todo{Notation for partially applied functor used in \figref{default-semantics:types}}

>>>>

% \input{fig/positivity} -- follow nunes2023 instead
\input{fig/syntax}
% \input{fig/kinding}
\input{fig/typing}

@@

% \input{fig/mu-polynomial}
\input{fig/semantics-parameterised-types}
\input{fig/semantics}

% \begin{definition}[$\mu$-polynomial]
% Suppose $\cat{C}$ a category with finite coproducts $(\coprod, 0)$ and finite products $(\times, 1)$. Define
% $\muPoly_{\cat{C}}$ to be the smallest subcategory of $\Cat$ generated inductively by the rules in
% \figref{mu-polynomial}.
% \end{definition}
%
% For any $\mu$-polynomial endofunctor $F: \cat{C} \to \cat{C}$ with initial algebra $\mu F$, write $\inMap_F$
% for the structure map $F\mu F \to \mu F$ and write $\cata_f$ for the unique morphism $\mu F \to X$ from the
% initial $F$-algebra to any $F$-algebra $(X, f: FX \to X)$.

To give the semantics for the language defined in \figrefTwo{syntax}{typing}, we fix a bicartesian closed
category $\Sem$ with finite products $(\times, 1)$, finite coproducts $(\coprod, 0)$, exponentials
$\internalHom{X}{Y}$, evaluation morphisms $\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$, plus the
following additional structure:
\begin{enumerate}
\item a strong monad $\mathcal{L}$ (lifting) with unit $\eta_X: X \to \mathcal{L}(X)$, multiplication $\mu_X:
\mathcal{L}^2(X) \to \mathcal{L}(X)$ and strength $\mathsf{st}_{X,Y}: X \times \mathcal{L}(Y) \to
\mathcal{L}(X \times Y)$;
% \item $\mu$-polynomial endofunctors given by $\muPoly_{\Sem}$;
\item for each primitive type $\rho \in \PrimTy$, an object $\sem{\rho}$;
\item for each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\phi}:
\sem{\rho_1} \times \ldots \times \sem{\rho_n} \to \sem{\rho}$.
\end{enumerate}

\figref{semantics:types} gives the interpretation of types $\Delta \vdash \tau$ with $\length{\Delta}
= n$ as functors $\Sem^n \to \Sem$, or as objects of $\Sem$ in the case $n = 0$.
\figref{semantics:terms} gives the interpretation of terms $\Gamma \vdash t: \tau$ as morphisms
$\sem{\Gamma} \to \sem{\tau}$.

% \todo{Notation for partially applied functor used in \figref{semantics:types}}

### 2025-06-28 notes/language.tex
Separate figure for parameterised types.

\input{fig/mu-polynomial}
\input{fig/default-semantics}

@@

\figref{default-semantics:types} gives the interpretation of types $\Delta \vdash \tau$ with $\length{\Delta}
= n$ as functors $\Sem^n \to \Sem$, or as objects of $\Sem$ in the case $n = 0$.
\figref{default-semantics:terms} gives the interpretation of terms $\Gamma \vdash t: \tau$ as morphisms
$\sem{\Gamma} \to \sem{\tau}$.

\todo{Notation for partially applied functor used in \figref{default-semantics:types}}

>>>>

\input{fig/mu-polynomial}
\input{fig/semantics}

@@

\figref{semantics:types} gives the interpretation of types $\Delta \vdash \tau$ with $\length{\Delta}
= n$ as functors $\Sem^n \to \Sem$, or as objects of $\Sem$ in the case $n = 0$.
\figref{semantics:terms} gives the interpretation of terms $\Gamma \vdash t: \tau$ as morphisms
$\sem{\Gamma} \to \sem{\tau}$.

\todo{Notation for partially applied functor used in \figref{semantics:types}}
